Placing the spotted T Tauri star LkCa 4 on an HR diagram

### 2016-09-18 document/ms.tex
Closes #44 : Feedback from the ASASSN team.  Still need to format the coauthors and affiliations though.

\author{Michael A. Gully-Santiago,\altaffilmark{1} Gregory J. Herczeg,\altaffilmark{1} Ian Czekala,\altaffilmark{2}, Garrett Somers\altaffilmark{2}, et al.}

\altaffiltext{1}{Kavli Institute for Astronomy and Astrophysics, Beijing, China}
\altaffiltext{2}{TBD}

\begin{abstract}
Magnetic fields temporarily increase the opacity in stellar photospheres, impacting both their instantaneous stellar flux and the duration and efficiency of their convective energy transport during pre-main sequence stellar evolution.  The observational manifestation of local magnetic fields, starspots on young stars can cover a large fraction of the stellar surface, causing the incorrect assessment of both the effective temperature and the luminosity and therefore the placement of sources on HR diagrams.  The singleton weak-lined T-Tauri star LkCa 4 shows exceptionally large photometric modulation attributable to star spots, making it an ideal testbed for measuring starspot properties.  Infrared spectral features in high resolution IGRINS spectra indicate that 80\% of the surface is covered in $\teffb \sim 2750$ K patches of cool photosphere at the epoch of the observation.  The rest of the $20\%$ of the stellar surface can be thought of as either ambient photosphere or a hot spot of $\teffa \sim4100$ K.  The spectral energy distribution, color variability, rotational modulation, and low-resolution optical/IR spectroscopy-- when registered to the filling factors at the epochs of their observations--strengthen the case for a large starspot filling factor and temperature contrast through time.  The revised effective temperature and luminosity make LkCa 4 appear much lower mass and much younger when placed on a pre-main sequence HR diagram, compared to previous estimates based only on optical spectroscopy and not taking into account the large coverage fraction of cool spots.  The derived starspot properties explain the extreme $V-$band photometric amplitude of 0.5$-$0.8 magnitudes.  Geometric considerations of spot coverage suggest that other heavily spotted young stars have been misplaced on the observational HR diagram, leading to incorrect estimates for ages and masses and contributing, in part, to the apparent age spread observed towards young clusters.
\end{abstract}

@@

The causes of global age uncertainties and of large luminosity spreads in individual clusters are controversial.  Observationally, \citet{hartmann01} and \citet{slesnick08} argue that measurement uncertainties may mask any real differences in ages within a cluster.  In models, \citet{hartmann97} and \citet{baraffe09} describe how non-static accretion histories change the stellar contraction at early times.  Contraction rates also depend on the prescription for convection, which may vary with mass.

Magnetic activity is a likely source of significant uncertainty in both the models and observations of young low-mass stars. Convection at these young ages generates strong magnetic fields, as measured in Zeeman broadening and polarimetry \citep[e.g.][]{johnskrull07,donati09} and as seen in starspots \citep[e.g.][]{stauffer03,grankin08}.  Evolutionary models are just now starting to implement new prescriptions for convection with magnetic fields (\citet{somers15}, \citet{feiden16}; see also \citet{baraffe15} for an updated treatment of convection without introducing magnetic fields).  Stellar evolution models including the effect of starspots can make a coeval 10 Myr population exhibit apparent age spreads of 3$-$10 Myr, with derived masses biased towards lower masses \citep{somers15}.  Starspots may also be responsible for biases in stellar effective temperatures derived by different methods.  For example, effective temperatures for 3493 young stars measured using the APOGEE spectrograph \citep[$1.5-1.70 \;\mu$m at $R=22,500$][]{wilson10} are offset by 200$-$500 K ($\sim$0.02$-$0.05 dex) and as high as 1000 K ($\sim$0.1 dex) relative to previous measurements, usually from optical \citep{cottaar14}.

@@

Starspots have recently been observed with exceptional photometric precision from monitoring surveys targeting exoplanet transits \citep[\emph{e.g}][]{harrison11,davenport15}.  Monochromatic photometric monitoring cannot separate the relative areas and temperatures of the emitting regions, and therefore cannot yield estimates for the areal fill factor of spots without making assumptions about the relative temperature contrast. Contemporaneous or near-contemporaneous panchromatic photometric monitoring \citep{petrov94,bouvier95,grankin07,cody14} can break the degeneracy between starspot areal filling factor and starspot temperature, but these measurements are still limited to detecting large longitudinally asymmetric starspots.  Transiting planets or eclipsing binaries passing in front of starspots \citep{desert11} can break geometric degeneracies, providing estimates for the characteristic sizes and lifetimes of starspots or starspot groups, but the number of such transiting systems and demands of high photometric precision limit wider application.  Zeeman Doppler Imaging \citep[ZDI]{donati14} suffers from similar longitudinal asymmetry limitations as photometry, but can yield some information on the latitudinal distribution of starspots in reconstructed brightness maps based on models for spectro-polarimetric line shapes.  All of these techniques would tend to underestimate the starspot areal coverage of pole-on stars, stars with longitudinally symmetric bands of starspots, or stars with an isotropic distribution of small starspots.

@@

Figure \ref{fig:PhotPhase} shows all available $V-$band data grouped by the 22 observing seasons and phase-folded by period $P=3.375$ days obtained from multiterm Lomb-Scargle periodograms \citep{ivezic14}.  The general appearance of the phase-folded lightcurves does not change with perturbations to the period on the scale of 0.003 days.  The estimated $BVR$ magnitudes at the time of the spectral observations are determined from regularized multiterm fits \citep{vanderplas15a} shown as the solid blue lines in Figure \ref{fig:PhotPhase}.  Table \ref{tbl_estimated_V} lists the estimated $BVR$ photometry and the observing epoch for data from IGRINS, ESPaDOnS, 2MASS \citep{skrutskie06}, DoubleSpec, and TripleSpec.

\begin{figure*}
 \centering
 \includegraphics[width=0.95\textwidth]{figures/Vband_22s.pdf}
 \caption{Phase-folded lightcurves constructed assuming the same period for all observing seasons.  The blue solid lines show a multi-term regularized periodic fit keeping the first $M_{\rm max}=4$ Fourier components.  The vertical lines show the epochs of observations for ancillary data, with the same line styles and colors as Figure \ref{fig:PhotTime}.  The unchanged vertical and horizontal scales highlight the secular drift of the light curve amplitude and morphology.}
 \label{fig:PhotPhase}
\end{figure*}

@@

Starspot line depth ratio analysis has traditionally been limited to isolated portions of spectrum that possess spectral lines attributable only to starspots and spectral lines attributable only to ambient photosphere \citep[\emph{e.g.}][]{neff95, oneal01}.  The apparent veiling of the lines and their respective temperature dependences can be combined to solve for the starspot and ambient photosphere temperatures and relative areal coverages.  The line depth ratio method suffers from the need to identify portions of the spectrum that possess such strong lines, and also suffers from the need to assemble large atlases of observed spectral templates to which the spotted star spectrum can be compared.  In this section, we introduce a generalization of line depth ratio analysis which employs pixel-by-pixel modeling of spectral orders.  By leveraging all the channels present in a spectrum, this strategy has the power to constrain starspot properties at relatively low fill-factors and identify weak lines that originate from the starspot.

We took two approaches to characterizing the effective temperature response from a spotted-star spectrum.  First we measured a single effective temperature to unique spectral orders in the optical (\S \ref{sec:ESP_starfish}) and near-IR (\S \ref{sec:IGR_starfish}).  Second we fit a two-temperature mixture model (\S \ref{sec:methods}) to a high grasp near-IR spectrum (\S \ref{sec:two_tempIGRINS}), which should show greater starspot flux density than optical.

@@

\citet[hereafter \iancze]{czekala15} developed a modular framework\footnote{The open source codebase and its full revision history is available at \url{https://github.com/iancze/Starfish}.  The experimental fork discussed in this paper is at \url{https://github.com/gully/Starfish}} to infer stellar properties from high resolution spectra.  The \iancze\ technique forward models observed spectra with synthetic spectral from pre-computed model grids.  The intra-grid-point spectra are emulated, taking into account the information loss attributable to the coarsely sampled stellar intrinsic parameters.  The forward model includes calibration parameters, line spread functions, and a Gaussian process noise model to account for correlations in the residual spectrum.  We employed the \PHOENIX\ grid of pre-computed synthetic stellar spectra, which span a wide range of wavelengths at high spectral resolution with sampling of 100 K in $\teff$ in our range of interest \citep{husser13}.  The modular framework was altered to accommodate starspot measurements in two ways. First, the single photospheric component was updated to include a starspot spectrum. Second, the MCMC sampling strategy was altered to accommodate the additional free parameters added in the starspot model.

The stellar photosphere is characterized as two photospheric components with a starspot temperature $\teffb$ and ambient photosphere temperature $\teffa$, with scalar solid angular coverages $\Omega_{\mathrm{spot}}$ and $\Omega_{\mathrm{amb}}$, respectively.    Starspots (or spot groups) are assumed to be cooler than the ambient photosphere, and they share the same average intrinsic and extrinsic stellar parameters otherwise; that is $(\vsini, \logg, \Z, v_z)$ are identical for starspots and ambient photosphere.  The composite mixture model for observed flux density can be written down as:

\begin{eqnarray} \label{eqn:mix_M}
S_{\mathrm{mix}} = \Omega_{\mathrm{amb}} B(\teffa)  + \Omega_{\mathrm{spot}} B(\teffb)
\end{eqnarray}

where $B$ is the spectral radiance from emulated model spectra.  The observed filling factor of starspots $f_{\Omega}$ is simply the ratio:

\begin{eqnarray} \label{eqn:fill_factor}
f_{\Omega} = \frac{\Omega_{\mathrm{spot}}}{\Omega_{\mathrm{amb}} + \Omega_{\mathrm{spot}}}
\end{eqnarray}

@@

MG-S and GJH are supported by general grant 11473005 awarded by the National Science Foundation of China.   The ESPaDOnS observations are supported by the contribution to the MATYSSE Large Project on CFHT obtained through the Telescope Access Program (TAP), which has been funded by the ``the Strategic Priority Research Program---The Emergence of Cosmological Structures'' of the Chinese Academy of Sciences (Grant No.11 XDB09000000) and the Special Fund for Astronomy from the Ministry of Finance.

>>>>

\author{Michael A. Gully-Santiago,\altaffilmark{1} Gregory J. Herczeg,\altaffilmark{1} Ian Czekala,\altaffilmark{2}, Garrett Somers\altaffilmark{3}, et al.}

%%% Coauthors:

%%% Konstantin Grankin
%CrAO

%%% Donati
% CNRS

%%% Covey
% WWU

%%% Greg Mace
%UT Austin

%%% TODO: Fill in more here as I go through the emails and feedback

%%%T.~W.-S.~Holoien\altafifilmark{3}
%- Department of Astronomy, The Ohio State University, 140 West 18th Avenue, Columbus, OH 43210, USA
%- Center for Cosmology and AstroParticle Physics (CCAPP), The Ohio State University, 191 W. Woodruff Ave., Columbus, OH 43210, USA

%%% Ben Shappee
%Hubble, Carnegie-Princeton Fellow
%Observatories of the Carnegie Institution for Science
%813 Santa Barbara Street
%Pasadena, CA  91101
%USA

\altaffiltext{1}{Kavli Institute for Astronomy and Astrophysics, Beijing, China}
\altaffiltext{2}{Stanford University, Palo Alto, CA}
\altaffiltext{3}{Department of Physics and Astronomy, Vanderbilt University, 6301 Stevenson Center, Nashville, TN, 37235, USA}

\begin{abstract}
Magnetic fields temporarily increase the opacity in stellar photospheres, impacting both their instantaneous stellar flux and the duration and efficiency of their convective energy transport during pre-main sequence stellar evolution.  The observational manifestation of local magnetic fields, starspots can cover a large fraction of the stellar surface, biasing estimates of both the effective temperature and the luminosity and therefore the placement of sources on HR diagrams.  The weak-lined T-Tauri star LkCa 4 shows exceptionally large photometric modulations of 0.5-0.8 $\Delta V$ mag attributable to star spots, making it an ideal testbed for measuring starspot properties.  Infrared spectral features in high resolution IGRINS spectra indicate that 80\% of the surface is covered in $\teffb \sim 2750$ K patches of cool photosphere at the epoch of the observation.  The rest of the $20\%$ of the stellar surface can be thought of as either ambient photosphere or a hot spot of $\teffa \sim4100$ K.  The spectral energy distribution, color variability, rotational modulation, and low-resolution optical/IR spectroscopy-- when registered to the filling factors at the epochs of their observations--strengthen the case for a large starspot filling factor and temperature contrast through time.  The revised effective temperature and luminosity make LkCa 4 appear much lower mass and much younger compared to previous estimates based only on optical spectroscopy and not taking into account the large covering fraction of the cool spots.  The derived starspot properties explain the extreme $V-$band photometric variability.  Emission from polar starspots viewed at moderate inclination has likely caused the misplacement of heavily spotted young stars on the observational HR diagram, leading to incorrect estimates for ages and masses and contributing, in part, to the apparent age spread observed towards young clusters.
\end{abstract}

@@

The causes of global age uncertainties and of large luminosity spreads in individual clusters are controversial.  Observationally, \citet{hartmann01} and \citet{slesnick08} argue that measurement uncertainties may mask any real differences in ages within a cluster.  In models, \citet{hartmann97} and \citet{baraffe09} describe how variable accretion histories change the stellar contraction and hence luminosities at early times.  Contraction rates also depend on the prescription for convection, which may vary with mass.

Magnetic activity is a likely source of significant uncertainty in both the models and observations of young low-mass stars. Convection at these young ages generates strong magnetic fields, as measured by Zeeman broadening and polarimetry \citep[e.g.][]{johnskrull07,donati09} and as seen in starspots \citep[e.g.][]{stauffer03,grankin08}.  Evolutionary models are just now starting to implement new prescriptions for convection with magnetic fields (\citet{somers15}, \citet{feiden16}; see also \citet{baraffe15} for an updated treatment of convection without introducing magnetic fields).  Stellar evolution models including the effect of starspots can make a coeval 10 Myr population exhibit apparent age spreads of 3$-$10 Myr, with standard mass estimates being biased towards lower masses \citep{somers15}.  Starspots may also be responsible for biases in stellar effective temperatures derived by different methods.  For example, the effective temperatures for 3493 young stars measured using the APOGEE spectrograph \citep[$1.5-1.70 \;\mu$m at $R=22,500$][]{wilson10} are offset by 200$-$500 K ($\sim$0.02$-$0.05 dex) and as high as 1000 K ($\sim$0.1 dex) relative to previous, usually optical, measurements \citep{cottaar14}.

@@

Starspots have recently been observed with exceptional photometric precision from monitoring surveys targeting exoplanet transits \citep[\emph{e.g}][]{harrison11,davenport15}.  The single band light curves from such surveys cannot separate the relative areas and temperatures of the emitting regions, and therefore cannot yield estimates for the filling factor of spots without making assumptions about the relative temperature contrast. Contemporaneous or near-contemporaneous panchromatic photometric monitoring \citep{petrov94,bouvier95,grankin07,cody14} can break the degeneracy between starspot filling factor and starspot temperature but these measurements are still limited to detecting large longitudinally asymmetric starspots.  Transiting planets or eclipsing binaries passing in front of starspots \citep{desert11} can break geometric degeneracies, providing estimates for the characteristic sizes and lifetimes of starspots or starspot groups, but the number of such transiting systems and demands of high photometric precision limit wider application.  Zeeman Doppler Imaging \citep[ZDI]{donati14} suffers from similar longitudinal asymmetry limitations as photometry, but can yield some information on the latitudinal distribution of starspots in reconstructed brightness maps based on models for spectro-polarimetric line shapes.  All of these techniques would tend to underestimate the starspot areal coverage of pole-on stars, stars with longitudinally symmetric bands of starspots, or stars with an isotropic distribution of small starspots.

@@

Figure \ref{fig:PhotPhase} shows all available $V-$band data grouped by the 22 observing seasons and phase-folded by the period $P=3.375$ days obtained from multiterm Lomb-Scargle periodograms \citep{ivezic14}.  The general appearance of the phase-folded lightcurves does not change with perturbations to the period on the scale of 0.003 days.  The estimated $BVR$ magnitudes at the time of the spectral observations are determined from regularized multiterm fits \citep[\emph{i.e.} Fourier series truncated to the first $\sim 4$ components]{vanderplas15a} shown as the solid blue lines in Figure \ref{fig:PhotPhase}.  Table \ref{tbl_estimated_V} lists the estimated $BVR$ photometry and the observing epoch for data from IGRINS, ESPaDOnS, 2MASS \citep{skrutskie06}, DoubleSpec, and TripleSpec.

\begin{figure*}
 \centering
 \includegraphics[width=0.95\textwidth]{figures/Vband_22s.pdf}
 \caption{Phase-folded lightcurves constructed assuming $P=3.375$ days for all 22 observing seasons.  The blue solid lines show a ``multiterm'' regularized periodic fit, that is, keeping the first $M_{\rm max}=4$ Fourier components \citep{vanderplas15a}.  The vertical lines show the epochs at which spectra or ancillary photometry were obtained, with the same line styles and colors as Figure \ref{fig:PhotTime}.  LkCa 4 shows secular changess in its light curve morphology.}
 \label{fig:PhotPhase}
\end{figure*}

@@

Starspot line depth ratio analysis has traditionally been limited to isolated portions of spectrum that possess spectral lines attributable only to starspots and spectral lines attributable only to ambient photosphere \citep[\emph{e.g.}][]{neff95, oneal01}.  The apparent veiling of the lines and their respective temperature dependences can be combined to solve for the starspot and ambient photosphere temperatures and relative areal coverages.  This line depth ratio method suffers from the need to identify portions of the spectrum that possess such strong lines, and from the need to assemble large atlases of observed spectral templates to which the spotted star spectrum can be compared.  In this section, we introduce a generalization of the line depth ratio analysis which employs pixel-by-pixel modeling of all spectral orders.  By using all the spectral data, this strategy has the power to constrain starspot properties at relatively low filling factors and to identify weak lines that originate from the starspots.

We took two approaches to characterizing the effective temperature response from a spotted-star spectrum.  First we measured a single unique effective temperature to each spectral order in the optical (\S \ref{sec:ESP_starfish}) and near-IR (\S \ref{sec:IGR_starfish}).  Second we fit a two-temperature mixture model (\S \ref{sec:methods}) to a near-IR echelle spectrum (\S \ref{sec:two_tempIGRINS}), which should show greater starspot flux density than optical.

@@

\citet[hereafter \iancze]{czekala15} developed a modular framework\footnote{The open source codebase and its full revision history is available at \url{https://github.com/iancze/Starfish}.  The experimental fork discussed in this paper is at \url{https://github.com/gully/Starfish}} to infer stellar properties from high resolution spectra.  The \iancze\ technique forward models observed spectra with synthetic spectra from pre-computed model grids.  The intra-grid-point spectra are ``emulated'' in a process similar but superior to interpolation, since it seemlessly quantifies the uncertainty attributable to the coarsely sampled stellar intrinsic parameters (see Appendix of \iancze).  The forward model includes calibration parameters, line spread functions, and a Gaussian process noise model to account for correlations in the residual spectrum.  We employed the \PHOENIX\ grid of pre-computed synthetic stellar spectra, which span a wide range of wavelengths at high spectral resolution with a sampling of 100 K in $\teff$ in our range of interest \citep{husser13}.  The modular framework was altered to accommodate starspot measurements in two ways. First, the single photospheric component was updated to include a starspot spectrum. Second, the MCMC sampling strategy was altered to accommodate the additional free parameters added by the starspot model.

The stellar photosphere is characterized as two photospheric components with a starspot temperature $\teffb$ and ambient photosphere temperature $\teffa$, with scalar solid angular coverages $\Omega_{\mathrm{spot}}$ and $\Omega_{\mathrm{amb}}$, respectively\footnote{For flux calibrated spectra, the total solid angle $\Omega$ can be constrained, but for typical echelle spectrographs only relative values of $\Omega_{\mathrm{spot}}$ and $\Omega_{\mathrm{amb}}$ can be inferred.}.  Starspots (or spot groups) are assumed to be cooler than the ambient photosphere, but otherwise share the same average intrinsic and extrinsic stellar parameters $(\vsini, \logg, \Z, v_z)$.  The composite mixture model for observed flux density is:
\begin{eqnarray} \label{eqn:mix_M}
S_{\mathrm{mix}} = \Omega_{\mathrm{amb}} B(\teffa)  + \Omega_{\mathrm{spot}} B(\teffb)
\end{eqnarray}
where $B(T)$ is the spectral radiance from model spectra.  The filling factor of the starspots is:
\begin{eqnarray} \label{eqn:fill_factor}
f_{\Omega} = \frac{\Omega_{\mathrm{spot}}}{\Omega_{\mathrm{amb}} + \Omega_{\mathrm{spot}}}
\end{eqnarray}

@@

MG-S and GJH are supported by general grant 11473005 awarded by the National Science Foundation of China.   The ESPaDOnS observations are supported by the contribution to the MATYSSE Large Project on CFHT obtained through the Telescope Access Program (TAP), which has been funded by the ``the Strategic Priority Research Program---The Emergence of Cosmological Structures'' of the Chinese Academy of Sciences (Grant No.11 XDB09000000) and the Special Fund for Astronomy from the Ministry of Finance. 
TW-SH is supported by the DOE Computational Science Graduate Fellowship, grant number DE-FG02-97ER25308.
